Parametricity Consequences of GADTs Representations

### 2021-03-01 20-gadts/GADT-story.tex
Work on Section 5 up to Example 5.4

To compute the interpretation of $\mathsf{G}$'s primitive
representation we first note that the type of $\mathsf{G}$'s solitary
constructor $\mathsf{c :: G\,1}$ can be equivalently expressed as
$\mathsf{c :: 1 \to G\,1}$ or, using~\eqref{eq:nat-transfs}, as
$\mathsf{c :: (Lan_{\lambda *. 1}\,\lambda *. 1)\,a \to G\, a}$, where
$\mathsf{\lambda *. 1}$ is the syntactic reflection of the constantly
$1$-valued functor from the category with a single object $*$ to
$\mathit{Set}$. We can therefore represent $\mathsf{G}$ as
\[\mathsf{G\,a} = \mathsf{(\mu \phi. \lambda b.
 (Lan_{\lambda *. 1} \,\lambda *.  1)\,b)\,a}\] The interpretation of
$\mathsf{G}$ is therefore the application to $\mathsf{a}$ of the
fixpoint of the interpretation of the body $\mathsf{\lambda
  b. (Lan_{\lambda *. 1} \,\lambda *. 1)\,b}$ of the syntactic
fixpoint $\mathsf{\mu \phi. \lambda b.  (Lan_{\lambda *. 1} \,\lambda
  *.  1)\,b}$. But since the recursion variable $\mathsf{\phi}$ does
not appear in this body, the interpretation of the fixpoint is just
the interpretation of the body itself.  The interpretation of
$\mathsf{G}$ is therefore $(\mathit{Lan}_{\lambda *. 1} \lambda
*. 1)\, A$, where $A$ interprets $\mathsf{a}$.
%and $1$ is the {\color{blue} constantly $1$-valued
%  functor $\ast \to \set$ from the category with a single object
%  $\ast$ to $\set$.}
It turns out, however, that, for any set $A$, $(\mathit{Lan}_{\lambda
  *. 1} \lambda *. 1)\,A$ is, in fact, exactly $A$. Indeed,
Proposition~7.1 of~\cite{blw03} gives that $(\mathit{Lan}_{\lambda
  *. 1} \lambda *. 1)\,A$ can be computed as the set of all functions
$a : 1 \to A$ modulo the smallest equivalence relation $\sim$ such
that, for any $f : 1 \to 1$, we have $a \sim a \circ f$.  But this set
is isomorphic to $A$, because the only function $f : 1 \to 1$ is the
identity function.

Putting it all together, we see that the Church encoding of
$\mathsf{G}$ and its primitive representation are not equivalent: the
interpretation of the former has exactly one data element at instance
$\mathsf{G\,1}$ and no elements at any other instances, whereas the
interpretation of the latter has data elements at every instance other
than $\mathsf{G\,0}$.
%Indeed,
These additional data elements can be obtained by reflecting back into
syntax the elements $\mathit{map}_G\,f_a\,c : G\,A$ resulting from
applying the functorial action $\mathit{map}_G$ of $\mathsf{G}$'s
interpretation $G$ to the functions $f_a : 1 \to A$ determined by the
elements $a$ of $A \not = \emptyset$ and the interpretation $c$ of
$\mathsf{c}$.
\end{example}

@@

The reader familiar with short cut fusion will quickly see that the
non-uniform nature of GADTs that are not nested types entails that
short cut fusion cannot be validated for their Church encodings. In
fact, it is not even possible to define a $\mathsf{build}$ function,
which is a crucial ingredient in short cut fusion, for the Church
encoding of a GADT that is not a nested type. But the fact that not
all of the elements in the functorial completion of a GADT that is not
a nested type are directly representable as applications of data
constructors means that it is also impossible to define a
$\mathsf{build}$ function for the primitive representation of such a
GADT.  This is not asurorising, however: if short cut fusion were to
hold for a GADT then its Church encoding and primitive representation
would necessarily be the equivalent.

In addition to not being equivalent, the two representations of GADTs
have very different implications for parametricity for GADTs. We
explore the differences in parametricity results for the two
representations in the next section.

\section{Parametricity in the Presence of GADTs}\label{sec:par}

{\em Relational parametricity} encodes a powerful notion of
type-uniformity, or representation independence, for data types in
functional languages. It formalizes the intuition that a polymorphic
program must act uniformly on all of its possible type instantiations
by requiring that every such program preserves all relations between
pairs of types at which it is instantiated. Parametricity was
originally put forth by Reynolds~\cite{rey83} for System F. It was
later popularized as Wadler's ``theorems for free''~\cite{wad89}, so
called because it can deduce properties of programs in such languages
solely from their types, i.e., with no knowledge whatsoever of the
text of the programs involved.  Most of Wadler's free theorems are
consequences of naturality for polymorphic list-processing
functions. However, parametricity can also derive results that go
beyond just naturality, such as establishing {\color{blue} proving}
the equivalence of the Church encodings and the primitive
representations of ADTs and {\color{blue} hereditarily polynomial}
nested types by validating the aforementioned short cut fusion
transformation for them.

{\color{blue} Say somewhere: A relation is a triple $(A,B,R)$, where
  $R$ is a relation between sets $A$ and $B$.  A function from
  $(A,B,R)$ to $(A',B',R')$ is a pair $(f : A \to A',g : B \to B')$ of
  functions on sets such that $(f a,g\,b) \in R'$ if $(a,b) \in R$. We
  may write $R : \rel(A,B)$ for $(A,B,R)$.  If $R : \rel(A,B)$ we
  write $\pi_1 R$ and $\pi_2 R$ for the {\em domain} $A$ of $R$ and
  the {\em codomain} $B$ of $R$, respectively, and assume $\pi_1$ and
  $\pi_2$ are surjective. We write $\Eq_A = (A,A,\{(x,x)~|~ x \in
  A\})$ for the {\em equality relation} on the set $A$. ALSO, WANT OUR
RELATIONS TO BE ONTO.}

{\color{red} This might not be relevant, since we only care about the
  fibredness.} 
The key idea underlying Reynolds' parametricity is to give each type
$F(\alpha)$ with one free variable $\alpha$ a {\em set interpretation}
$F_0$ taking sets to sets and a \emph{relational interpretation} $F_1$
taking relations $R : \rel(A,B)$ to relations $F_1 (R) : \rel(F_0 (A),
F_0 (B))$, and to interpret each term $t(\alpha,x) : F(\alpha)$ with
one free term variable $x : G(\alpha)$ as a map $t_0$ associating to
each set $A$ a function $t_0(A) : G_0(A) \to F_0(A)$. These
interpretations are given inductively on the structures of $F$ and $t$
in such a way that they imply two fundamental theorems. The first is
an \emph{Identity Extension Lemma}, which states that $F_1(\Eq_A) =
\Eq_{F_0(A)}$, and is the essential property that makes a model
relationally parametric rather than just induced by a logical
relation. The second is an \emph{Abstraction Theorem}, which states
that, for any $R :\rel(A, B)$, $(t_0(A),t_0(B))$ is a morphism in
$\rel$ from $(G_0(A),G_0(B),G_1(R))$ to $(F_0(A),F_0(B),F_1(R))$. The
Identity Extension Lemma is similar to the Abstraction Theorem except
that it holds for {\em all} elements of a type's interpretation, not
just those that interpret terms.  Similar theorems are required for
types and terms with any number of free variables.

\begin{example}
{\color{red} PRIMITIVE REP DOESN'T HAVE FIBRED SEMANTICS, AS REQUIRED
  FOR PAR} To
even consider the question of whether our semantics is parametric we
first need to define a relational semantics for our types, i.e., a
relational interpretation for every type $G : \ast \to \ast$ such
that, for every relation $R : \rel(A, B)$, we have $G_1(R) :
\rel(G_0(A), G_0(B))$.

We shall show that this is not generally possible for the primitive
representation of data types, by considering the previously defined
type $G$ in Equation~\ref{eq:G}.  Recall that the semantic
interpretation of $G$ is given by $\mathsf{\Lan}_1 1$ which, in fact,
is none other than the identity functor on $\set$.

Analogously to the set interpretation, the relational interpretation
for $G$ is given by $\mathsf{\Lan}_{\Eq_1} \Eq_1$.  Notice that the
interpretation still essentially relies on a left Kan extension, but
this time it is an extension of $\rel$-valued functors.  We would need
to prove that, for every relation $R : \rel(A, B)$, we have that
$(\mathsf{\Lan}_{\Eq_1} \Eq_1) R$ is a relation on $(\mathsf{\Lan}_1
1) A$ and $(\mathsf{\Lan}_1 1)B$, i.e., $A$ and $B$.  Instead, we will
provide a counterexample showing that this is not true.

@@

We still get naturality properties, though.

@@

As we have seen in the previous Section~\ref{sec:adts}, GADTs
represented as Church encodings do not necessarily have a map or a
functorial semantics.  In particular, this means that we cannot
possibly derive naturality results for those types, event though their
semantics is parametric.  Indeed, even though naturality is sometimes
seen as a consequence of parametricity, that is only true in case we
have a functorial semantics and types supporting a map.  Still, GADTs
represented as Church encodings can still exhibit consequences of
paramentricity other than naturality, as they do admit parametric
semantics~\cite{atk12}.

\begin{example}
{\color{blue} In this paragraph, $\mathsf{G}$ and $\mathsf{t}$
  represent syntax, so should be sans-serif, but in the notes, the
  paragraph about set/relational interpretation and
  parametricity/IEL/AT uses serif font.} For example, consider the
Church encoding of the GADT $\mathsf{G}$ defined in
Equation~\ref{eq:G}.  If $\mathsf{t}$ is a term of type
$\mathsf{G}$ and $R$ is a relation on the sets $A$ and $B$, then
$t_0(A) \in G_0(A)$ and $t_0(B) \in G_0(B)$, and by the abstraction
theorem we have that $t_0(A)$ and $t_0(B)$ are related in $G_1(R)$.
{\color{blue} Here, $G = (G_0,G_1)$ is the interpretation of
  $\mathsf{G}$.}  We also know that $G_1(R)$ is the empty relation
whenever $R$ is not $\mathit{Equal}\,1$.  We thus deduce that there
can be no term of type $\mathsf{G\,a}$ unless $\mathsf{a}$ is
semantically equivalent to $\mathsf{1}$.  This result is the syntactic
counterpart of the inhabitation result for $\mathsf{G}$ {\color{blue}
  are inhabitation results about semantics or syntax?}  showing that
all of the instances of $\mathsf{G}$ are uninhabited, with the
exception of the instance for $\mathsf{1}$.  Of course, we could not
possibly have a term of an uninhabited type.
\end{example}

%%% Primitive representation of GADT is non parametric, but functorial
On the contrary, as we have shown in Section~\ref{sec:gadts}, a
GADT represented as a primitive data type does not necessarily admit a
parametric semantics, but, as a fixpoint of a higher-order functor, it
does admit a functorial semantics and, consequently, a functorial map.
In particular, every type that has a functorial map also benefits from
naturality properties.  In other words, there is a suitable
rearrange-transform property as explained in Section~\ref{sec:gadts}.

%%% Example with G
For example, $\mathsf{G}$ is associated to a functorial map,
$\map_{\mathsf{G}}$, such that, for every function $\mathsf{f : a \to
  b}$, we get a function $\map_{\mathsf{G}}\,\mathsf{f} : \mathsf{G\,a
  \to G\,b}$.  Then we have a term $\map_{\mathsf{G}}\,\mathsf{f\,c} :
\mathsf{G\,a}$ for every function $\mathsf{f : \mathsf{1} \to a}$,
where $\mathsf{c : G\,\mathsf{1}}$ is the constructor provided by the
definition of $\mathsf{G}$.

We can read our counterexample as showing that parametricity results
for calculi representing GADTs as their Church encodings cannot be
extended to calculi that represent GADTs as primitive data types. And
functoriality results for calucli representing GADTs as primitive
cannot be extended to calculi that represent GADTs as their Church
encodings.  So the user is forced to make a choice between naturality
consequences of parametricity and other consequences of
parametricity. If GADTs really did generalize ADTs, and really were
data types in a normal sense, then these would coincide and the user
would not have to choose.

\vspace*{1in}

>>>>

To compute the interpretation of $\mathsf{G}$'s primitive
representation we first note that the type of $\mathsf{G}$'s solitary
constructor $\mathsf{c :: G\,1}$ can be equivalently expressed as
$\mathsf{c :: 1 \to G\,1}$ or, using~\eqref{eq:nat-transfs}, as
$\mathsf{c :: (Lan_{\lambda *. 1}\,\lambda *. 1)\,a \to G\, a}$, where
$\mathsf{\lambda *. 1}$ is the syntactic reflection of the constantly
$1$-valued functor from the category with a single object $*$ to
$\mathit{Set}$. We can therefore represent $\mathsf{G}$ as
\[\mathsf{G\,a} = \mathsf{(\mu \phi. \lambda b.
 (Lan_{\lambda *. 1} \,\lambda *.  1)\,b)\,a}\] The interpretation of
$\mathsf{G}$ is therefore the application to $\mathsf{a}$ of the
fixpoint of the interpretation of the body $\mathsf{\lambda
  b. (Lan_{\lambda *. 1} \,\lambda *. 1)\,b}$ of the syntactic
fixpoint $\mathsf{\mu \phi. \lambda b.  (Lan_{\lambda *. 1} \,\lambda
  *.  1)\,b}$. But since the recursion variable $\mathsf{\phi}$ does
not appear in this body, the interpretation of the fixpoint is just
the interpretation of the body itself.  The interpretation of
$\mathsf{G\,a}$ is therefore $(\mathit{Lan}_{\lambda *. 1} \lambda
*. 1)\, A$, where $A$ interprets $\mathsf{a}$.
%and $1$ is the {\color{blue} constantly $1$-valued
%  functor $\ast \to \set$ from the category with a single object
%  $\ast$ to $\set$.}
It turns out, however, that, for any set $A$, $(\mathit{Lan}_{\lambda
  *. 1} \lambda *. 1)\,A$ is, in fact, exactly $A$. Indeed,
Proposition~7.1 of~\cite{blw03} gives that $(\mathit{Lan}_{\lambda
  *. 1} \lambda *. 1)\,A$ can be computed as the set of all functions
$a : 1 \to A$ modulo the smallest equivalence relation $\sim$ such
that, for any $f : 1 \to 1$, we have $a \sim a \circ f$.  But this set
is isomorphic to $A$, because the only function $f : 1 \to 1$ is the
identity function.

Putting it all together, we see that the Church encoding of
$\mathsf{G}$ and its primitive representation are not semantically
equivalent: the interpretation of the former has exactly one data
element at instance $\mathsf{G\,1}$ and no elements at any other
instances, whereas the interpretation of the latter has data elements
at every instance other than $\mathsf{G\,0}$.
%Indeed,
These additional data elements can be obtained by reflecting back into
syntax the elements $\mathit{map}_G\,f_a\,c : G\,A$ resulting from
applying the functorial action $\mathit{map}_G$ of $\mathsf{G}$'s
interpretation $G$ to the functions $f_a : 1 \to A$ determined by the
elements $a$ of $A \not = \emptyset$ and the interpretation $c$ of
$\mathsf{c}$.
\end{example}

@@

The reader familiar with short cut fusion will quickly see that the
non-uniform nature of GADTs that are not nested types entails that
short cut fusion cannot be validated for their Church encodings. In
fact, it is not even possible to define a $\mathsf{build}$ function,
which is a crucial ingredient in short cut fusion, for the Church
encoding of a GADT that is not a nested type. But the fact that not
all of the elements in the functorial completion of a GADT that is not
a nested type are directly representable as applications of data
constructors means that it is also impossible to define a
$\mathsf{build}$ function for the primitive representation of such a
GADT.  This is not surprising, however: if short cut fusion were to
hold for a GADT then its Church encoding and primitive representation
would necessarily be semantically equivalent.

In addition to not being semantically equivalent, the two
representations of GADTs have very different implications for
parametricity. We explore the differences in parametricity results for
GADTs with respect to the two representations in the next section.

\section{Parametricity for GADTs}\label{sec:par}

{\em Relational parametricity} encodes a powerful notion of
type-uniformity, or representation independence, for data types in
functional languages. It formalizes the intuition that a polymorphic
program must act uniformly on all of its possible type instantiations
by requiring that every such program preserves all relations between
pairs of types at which it is instantiated. Parametricity was
originally put forth by Reynolds~\cite{rey83} for System F. It was
later popularized as Wadler's ``theorems for free''~\cite{wad89}, so
called because it can deduce properties of programs in such languages
solely from their types, i.e., with no knowledge whatsoever of the
text of the programs involved.  Most of Wadler's free theorems are
consequences of naturality for polymorphic list-processing
functions. However, parametricity can also derive results that go
beyond just naturality, such as proving equivalence of Church
encodings and primitive representations of ADTs and {\color{blue}
  hereditarily polynomial} nested types by validating short cut fusion
for them.

To discuss relational parametricity will need to interpret data types
not just in $\mathit{Set}$, but in a suitable category of relations as
well. The following definition is standard:
\begin{definition}
  The category $\mathit{Rel}$ has:
  \begin{itemize}
\item objects: A relation is a triple $(A,B,R)$, where $R$ is a
  relation between sets $A$ and $B$.
\item morphisms: A morphism from $(A,B,R)$ to $(A',B',R')$ is a pair
  $(f : A \to A',g : B \to B')$ of functions in $\mathit{Set}$ such
  that $(f a,g\,b) \in R'$ if $(a,b) \in R$.
\item identities: The identity morphism on $(A,B,R)$ is the pair
  $(\mathit{id}_A : A \to A, \mathit{id}_B : B \to B)$.
\item composition: Composition is the componentwise composition in
  $\mathit{Set}$. 
\end{itemize}
\end{definition}
\noindent
We write $R : \mathit{Rel}\,(A,B)$ for $(A,B,R) \in \mathit{Rel}$.  If
$R : \mathit{Rel}\,(A,B)$ then we write $\pi_1 R$ and $\pi_2 R$ for
the {\em domain} $A$ of $R$ and the {\em codomain} $B$ of $R$,
respectively. Below, we assume $\pi_1$ and $\pi_2$ are surjective.  We
write $\mathit{Eq}_A = (A,A,\{(x,x)~|~ x \in A\})$ for the {\em
  equality relation} on the set $A$.

The key idea underlying parametricity is to give each type
$\mathsf{G[a]}$ with one free variable $\mathsf{a}$ a {\em set
  interpretation} $G_0$ taking sets to sets and a \emph{relational
  interpretation} $G_1$ taking relations $R : \mathit{Rel}\,(A,B)$ to
relations $G_1 \,R : \mathit{Rel}\,(G_0 \,A, G_0 \,B)$, and to
interpret each term $\mathsf{t\,(a,x) :: G[a]}$ with one free term
variable $\mathsf{x :: F[a]}$ as a function $t$ associating to each
set $A$ a morphism $t \,A : F_0\,A \to G_0\,A$ in $\mathit{Set}$.
These interpretations are given inductively on the structures of
$\mathsf{G}$ and $\mathsf{t}$ in such a way that they imply two
fundamental theorems. The first is an \emph{Identity Extension Lemma},
which states that $G_1\,\mathit{Eq}_A = \mathit{Eq}_{G_0 A}$, and is
the essential property that makes a model relationally parametric
rather than just induced by a logical relation.  The second is an
\emph{Abstraction Theorem}, which states that, for any $R
:\mathit{Rel}\,(A, B)$, $(t\, A, t\,B)$ is a morphism in
$\mathit{Rel}$ from $(F_0\,A,F_0\,B,F_1\,R)$ to
$(G_0\,A,G_0\,B,G_1\,R)$. The Identity Extension Lemma is similar to
the Abstraction Theorem except that it holds for {\em all} elements of
a type's interpretation, not just those that interpret terms.  Similar
theorems are required for types and terms with any number of free
variables.

The existence of parametric models that interpret GADTs as the
interpretations of their Church encodings follows from, e.g., the
existence of the parametric model of $F_\omega$ constructed
in~\cite{atk12}.  In that model, {\color{blue} ``parallel
  interpretations'' of $Rel$ over $Set$...}.  The parametricity
property for a GADT is therefore...  {\color{blue} SAY MORE!} In
particular, the parametricity property for the GADT $\mathsf{G}$
from~\eqref{eq:G} is exactly the syntactic counterpart of the
inhabitation result for the interpretation of $\mathsf{G}$'s Church
encoding, as the following example shows.

\begin{example}
Let $\mathsf{t}$ be a term of type $\mathsf{G}$ for the GADT
$\mathsf{G}$ defined in~\eqref{eq:G}, let $G = (G_0,G_1)$ be the
interpretation of the Church encoding of $\mathsf{G}$, let $t$ be the
interpretation of $\mathsf{t}$, and let $R :
\mathit{Rel}\,(A,B)$. Then $t\, A \in G_0\,A$ and $t\,B \in G_0\,B$
and, by the Abstraction Theorem, $t\,A$ and $t\,B$ are related in
$G_1\,R$. However, {\color{blue} using the semantic interpretations
  in~\cite{atk12} --- EXPLAIN!}, we have that the relational
interpretation $G_1\,R$ of $\mathsf{G}$ is the empty relation whenever
$R$ differs from the relational interpretation $\mathit{Eq}\,1$ of
$\mathsf{1}$.  {\color{blue} $Eq$ vs. $Equal$?}  We thus deduce that
there can be no term of type $\mathsf{G\,a}$ unless $\mathsf{a}$ is
semantically equivalent to $\mathsf{1}$.
\end{example}

For primitive representations of GADTs the story is completely
different. First off, we have that {\color{blue} the interp of Lan
  must be full Lan}. Indeed,... {\color{blue} MORE!!}
\begin{proposition}\label{prop:full-lans}
Full Lan prop goes here. Explain cutting down.
\end{proposition}
\noindent
In particular, both the set interpretation and the relational
interpetation of a GADT must involve the entireties of any Lans in
their fixpoint representations. We therefore have:

\begin{example}
In any parametric model we must give both a set interpretation and a
relational interpretion for every type as described at the start of
this section. That is, for every GADT $\mathsf{G}$ we must give an
interpretation $G = (G_0,G_1)$ such that, for every relation $R :
\mathit{Rel}\,(A, B)$, we have $G_1\,R : \mathit{Rel}\,(G_0\,A,
G_0\,B)$. But, as we now show, this is not generally possible for in
models that interpret GADTs as the interpretations of their primitive
representations.

Consider again the GADT $\mathsf{G}$ given by~\eqref{eq:G}.  Applying
Proposition~\ref{prop:full-lans} first in $\mathit{Set}$ and then in
$\mathit{Rel}$, we have that the set interpretation of $\mathsf{G}$ is
$\mathit{Lan}_{\lambda *.1}\,\lambda *.1$, i.e., is exactly the
identity functor on $\mathit{Set}$ {\color{blue} Tie back to previous
  calculation}. By the exact same reasoning, this time in
$\mathit{Rel}$ rather than in $\mathit{Set}$, the relational
interpretation of $\mathsf{G}$ is $\mathit{Lan}_{\lambda
  *.\mathit{Eq}_1} \,\lambda *\!.\,\mathit{Eq}_1$, where $\lambda
*\!.\,\mathit{Eq}_1$ is the constantly $\mathit{Eq}_1$-valued functor
from the category with a single object $*$ to $\mathit{Rel}$.  Indeed,
the interpretation is still a left Kan extension, but now it is the
left Kan extension determined by the functor interpreting the type
$\mathsf{\lambda *. 1}$ in $\mathit{Rel}$. To prove that the Identity
Extension Lemma holds we need to show that, for every relation $R :
\mathit{Rel}\,(A, B)$, we have that $(\mathit{\Lan}_{\lambda
  *\!.\,\mathit{Eq}_1} \,\lambda *\!.\,\mathit{Eq}_1)\, R$ is a
relation between the sets $(\mathit{Lan}_{\lambda *. 1}\,\lambda
*\!. 1)\, A$ and $(\mathit{Lan}_{\lambda *. 1}\,\lambda *\!. 1)\,B$,
i.e., between the sets $A$ and $B$.  Instead, we will give a
counterexample showing that this is not the case.

Consider the relation $R = (1, 2, 1 \times 2)$. We expect
$(\mathit{\Lan}_{\lambda *\!.\,\mathit{Eq}_1} \lambda
*\!.\,\mathit{Eq}_1)\, R$ to be a relation with domain {\color{blue}
  $1$}. Since left Kan extensions preserve projections~\cite{rie16},
we can compute the domain as
\[\pi_1\,\big( (\mathit{\Lan}_{\lambda *\!.\,\mathit{Eq}_1} \lambda
*\!.\,\mathit{Eq}_1) \, R \big)\;=\; (\mathsf{\Lan}_{\lambda
  *\!.\,\mathit{Eq}_1} \lambda *\!. \,1)\, 1\]

(notice that the left
Kan extension of $1 : \ast \to \set$ along $\Eq_1 : \ast \to \rel$ is
a functor $\rel \to \set$).

@@

@@

Although there can exist no parametric models interpreting GADTs as
their primitive representations, we still get naturality properties
for prim reps. On the other hand, although a GADT represented as a
Church encoding does not necessarily have a map or a functorial
semantics, so that we cannot possibly derive naturality results for
them, parametric models interpreting GADTs as their Church encodings
do exist.  This shows that, even though naturality is generally
regarded as a consequence of parametricity, this is only true when we
have a functorial semantics and types supporting a map.

%%% Primitive representation of GADT is non parametric, but functorial
%On the contrary, as we have shown in Section~\ref{sec:gadts}, a
%GADT represented as a primitive data type does not necessarily admit a
%parametric semantics, but, as a fixpoint of a higher-order functor, it
%does admit a functorial semantics and, consequently, a functorial map.
%In particular, every type that has a functorial map also benefits from
%naturality properties.  In other words, there is a suitable
%rearrange-transform property as explained in Section~\ref{sec:gadts}.

We can read the results of this section as showing that parametricity
results for calculi representing GADTs as their Church encodings
cannot be extended to calculi that represent GADTs as primitive data
types. And functoriality results for calucli representing GADTs as
primitive cannot be extended to calculi that represent GADTs as their
Church encodings.  So the user is forced to make a choice between
naturality consequences of parametricity and other consequences of
parametricity. If GADTs really did generalize ADTs, and really were
data types in a normal sense, then these would coincide and the user
would not have to choose.
